Process-based phenotypes: an application to the Cellular Phenotype Ontology

### 2011-11-25 paper/drinking.tex

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while a particular {\em goal line}
can participate in the same process in a different role.

@@

All GO regulation terms use one of three relations, regulates,
negatively_regulates and positively_regulates, to link regulation
terms to process or quality terms. The regulates relations are defined
in terms of qualities: a regulatory process causes a change in
magnitude to some quality, which in turn has an effect on the
frequency, rate or duration of some other type of process. Effects
that results in increases and decreases use positively_regulates and
negatively_regulates respectively. \cite{} (cite the paper on
cross-products in GO, PMID:20152934) The existing ontology structure
would also support the addition of subclasses to distinguish, for
example, regulation of the rate of a process from regulation of its
duration or time of onset.

@@

\subsection{PATO and the EQ model}

>>>>

Processes commonly have participants\footnote{It may be argued that
  processes necessarily have participants or that processes may have
  no participants. However, these discussions do not affect our aims
  here and take no position in this regard.}. Participants of
processes are entities that are present at time points (i.e., they
have no temporal parts) and may undergo changes during the process.
We can further distinguish different {\em roles} of participation in a
process. For example, a runner may participate in a {\em Triathlon}
process as a {\em Runner (role)}, while a {\em Referee} can
participate in the same process but in a different role.

@@

All GO regulation terms use one of three relations, regulates,
negatively\_regulates and positively\_regulates, to link regulation
terms to process or quality terms. The regulates relations are defined
in terms of qualities: a regulatory process causes a change in
magnitude to some quality, which in turn has an effect on the
frequency, rate or duration of some other type of process. Effects
that results in increases and decreases use positively\_regulates and
negatively\_regulates respectively. \cite{} (cite the paper on
cross-products in GO, PMID:20152934) The existing ontology structure
would also support the addition of subclasses to distinguish, for
example, regulation of the rate of a process from regulation of its
duration or time of onset.

@@

\subsection{PATO and the EQ model}
The Phenotype And Trait Ontology (PATO) was envisaged and designed to
provide a platform for allowing the integration of quantitative and
qualitative phenotype related information across different levels of
granularity of domain knowledge granularity and species
\cite{Gkoutos2005}.  PATO forms an ontology of phenotype qualities
that form basic entities that we can perceive and/or measure such as
colors, sizes, rates etc. One of it's classification axis lies on the
basis of the basic type of entity these qualities inhere to and
distinguishes between qualities of physical objects and qualities of
processes.

PATO allows for the description of affected entities by combining
various ontologies that describe the entities that have been affected,
such as the various anatomical ontologies, GO\cite {Ashburner2000b},
the Cell Type Ontology\cite {Bard2005a}, (SO)\cite {Eilbeck2005} etc
with the various qualities it provides for defining how these entities
were affected.  PATO can be used for annotation either directly in a
so called post-composed (post-coordinated) manner or for providing
formal (logical) definitions (equivalence axioms) to ontologies
containing a set of precomposed (pre-coordinated) phenotype terms. For
instance, to describe the decrease in the length of the sexual cycle
of female animals, we can combine the PATO term \textit{decreased
  duration} (PATO:0000499) with the Gene Ontology term \textit{estrous
  cycle} (GO:0042698), whilst if such a term existed in a precomposed
ontology (for the MP:0009007 term from the Mammalian Phenotype) it
could be used to provide an equivalence statement between that class
and the above PATO-based description.
